### 2022-03-30 manuscript.tex
csl [ci skip]
This build is based on
https://github.com/scherzerthesis/thesis/commit/1af40d40f8291eb85788543d7614044c788929f1.

This commit was created by the following CI build and job:
https://github.com/scherzerthesis/thesis/commit/1af40d40f8291eb85788543d7614044c788929f1/checks
https://github.com/scherzerthesis/thesis/actions/runs/2067060477

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/cacf67ddbb3ecfa8a81d823abdfbeafd930f1848/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/cacf67ddbb3ecfa8a81d823abdfbeafd930f1848}{scherzerthesis/thesis@cacf67d}
on March 30, 2022.

@@

\textbf{NSCLC} Lung cancer is a global hazard on human health. The historic consumption of cigarette smoke has caused lung cancer to be the most common and deadly form of cancer. Fortunately, smoking-rates have decreased in the past 30 years with lung cancer rates also on the decline. However, even if cigarette consumption were eradicated, lung cancer would still significantly affect human health. Indeedm it is estimated that approximately 25\% of all lung cancer cases arise in never smokers (Warren and Cummings, 2013). Secondary causes of lung cancer include Radon exposure, high-heat cooking, family history, air pollution (Malhotra \emph{et al.}, 2016). An individual's risk of lung cancer also increases with age due to decreases in the efficiency of tumor supressive mechanisms such as tumor immunity and apoptosis(de Groot and Munden, 2012).

Lung cancer can be broadly classified into two major histological categories that are named based on how the malignant cells look under a microscope: Non-small Cell Lung Cancer (NSCLC) and Small Cell Lung Cancer (SCLC). Within NSCLC, tumors are further characterized based on histological structures, anatomical location, cell-of-origin, as well as genetic drivers(Herbst \emph{et al.}, 2008). For example, lung adenocarcinoma is the most common subtype and is characterized by glandular and papillary structures while squamous cell lung carcinoma is characterized by keratin-pearls(Nikitin \emph{et al.}, 2004). Moreover, adenocarcinomas arise from alveolar-type-2 (AT2) cells in the distal lung and alveoli and are typified by genomic alterations throughout the MAPK pathway, such as EGFR, KRAS, BRAF, PI3K (Ferone \emph{et al.}, 2020; Jackson \emph{et al.}, 2001). On the other hand, squamous cells carcinomas likely arise from basal cells and typically harbor SOX2 gene amplification (Mukhopadhyay and Oliver, 2014; Mollaoglu \emph{et al.}, 2018).

The 5-year survival rate for patients with lung cancer was 25\% in 2020 but varies depending on stage of disease at time of diagnosis (Siegel \emph{et al.}, 2021).Like other solid cancers, lung cancers are most deadly after primary tumors have metastaized to other organ systems like the liver, kidney, bone, or nervous system (Matsunaga and Shida, 1971). Lung cancer patients with no local or distant metastasis typically have a 5-year survival percentage of approximately 40\% compared to approximately 5\% with distant metastasis{[}Woodard \emph{et al.}; 27740970{]}. Ultimately, the cause of death of many lung cancer patients include primary lung tumor burden, infection, pulmonary embolism or hemorrhage, or complications of metastatic dissemination(Nichols \emph{et al.}, 2012). Although patient survival has improved due to improved targeted and immune therapies, the still poor prognosis reflects a need to better understand the molecular mechanisms underlying lung adenocarcinoma initiation, progression, maintenance and response to targeted- or immune-therapy.

The current standard of care for Lung adenocarcinoma patients are often a combination of several conventional chemotherapeutics including Cisplatin, Carboplatin, Paclitaxel, or Pemetrexed(Ghimessy \emph{et al.}, 2020). Fortunately for patients who have been identified to have EGFR, ALK, ROS, TRK, or BRAF (V600E) genetic alternations are often offered pathway-targeted therapies(Planchard \emph{et al.}, 2016; Arteaga and Engelman, 2014). As of late, many patients receive immunotherapy first, followed by pathway-targeted therapy after progression on immunotherapy. Furthermore, in the last seven years, exciting therapeutic options for \emph{KRAS} mutant lung cancer have been developed and tested, such as a combination of MEK-inhibition and autophagy-inhibtion, and direct inhibitors of \emph{KRAS} G12c and \emph{KRAS} G12D (Kinsey \emph{et al.}, 2019; Reita \emph{et al.}, 2022; Zheng \emph{et al.}, 2022)

@@

The importance of this pathway is reflected in the strong anti-tumor responses that are seen when small molecule inhibitors are used to target various molecules involved in the MAPK pathway. For example, Dabrafenib and Trametinib, inhibitors of BRAF and MEK kinases, respectfully, are used in patients that harbor BRAF(V600E) melanomas and lung cancers(Planchard \emph{et al.}, 2021; Robert \emph{et al.}, 2014). Although there is strong pre-clinical and clinical evidence that BRAF+MEK blockade can lead to tumor regression, not all patients respond to treatment(Tabbò \emph{et al.}, 2021). Therefore, to see a complete response, multiple pathways need to be targeted, such as PI3K or autophagy (Deuker \emph{et al.}, 2014; Truong \emph{et al.}, 2020; Kinsey \emph{et al.}, 2019).

@@

\textbf{Genetically-engineered mouse models} Genetically-Engineered mouse models (GEMMs) of human cancer have allowed for the reproducible analysis of cancer biology driven by specific mutations frequently found in cancer patients. These models are helpful in understanding the molecular mechanisms that drive tumor formation and testing tumor responses to targeted or immune therapies. Early mouse models were not engineered but instead selected inbred mouse strains that were natural hosts of tumor transplantation(Strong, 1978). However, these models suffer from long latencies, incomplete pentrance, and often not relevant to human cancers. Therefore, carcinogen-induced models such as UV, Urethane, radiation, cigarette consumption, or MNU treatment models were developed to ressemble carinogene-induced human cancers(Kemp \emph{et al.}, 1994; Westcott \emph{et al.}, 2014).After techniques were developed to genetically manipulate embryonic stem-cells came the ability to engineer mice with specific mutations that could drive tumor formation. Since genes involved in cancer are also important for normal mouse development, conditional gene manipulation was developed to allow for temporal control.

The most commonly used conditional gene expression system relies on inducible transcription factors, such as tetracycline-dependent regulatory system, where the treatment of mice with doxycycline toggles gene transcriptial levels of transgenes. More recently, Cre-Lox recombination strategies have been utilized to get more cell type specific control. In this system, genes of interested are constructed to contain LoxP-sites that flank a particular genetic element and upon delivery of Cre-recombinase the DNA sequence in between LoxP sites are excised. With this approach, one can knockout a gene by engineering LoxP sites around critical exons, or one could induce gene transcription by flanking a strong stop signal such as a polyadenylation sequence upstream of a gene-of-interest sequence such as KRAS(G12D)(Jackson \emph{et al.}, 2001).

@@

Our lab has previously developed a conditional mouse model of BRAF (V600E) human lung cancers(Dankort \emph{et al.}, 2007; Shai \emph{et al.}, 2015; van Veen \emph{et al.}, 2019). In the \emph{Braf(CAT)} model, normal BRAF is expressed from a conditional allele prior to Cre-mediated recombination . LoxP sites flank human \emph{BRAF} cDNA encoding normal \emph{BRAF} exons 15-18. Downstream of the LoxP sites is the mutant exon 15 that encodes the murine equivalent of the T1799A mutation that gives rise to the BRAF (V600E) onco-protein. Downstream of the mutant exon there is a P2A element and a CAAX-tagged \emph{TdTomato}. Therefore, after Cre-mediated recombination, the BRAF (V600E) oncoprotein and fluorophore TdTomato is expressed at normal levels(van Veen \emph{et al.}, 2019).

\textbf{BRAF (V600E) Tumorigenesis and Cooperating Events for Progression} Mutational activation of the MAPK pathway forms lung tumors in mice. specifically, expression of BRAF (V600E) in the Surfactant-Protein C expression cells (alveolar type 2 pneumocytes) elicits benign tumor formation that fails to progress to lung adenocarcinoma. It is hypothesized that the cell cycle arrest is dependent on the tumor suppressors P53 and CDKN2A. Loss of either P53 or CDKN2A allows tumors to progress to malignant and deadly lung adenocarcinoma. Also, activation of WNT signaling and PI3K signaling similarly allows BRAF (V600E) adenomas to progress to adenocarcinoma (van Veen \emph{et al.}, 2019; Green \emph{et al.}, 2015; Juan \emph{et al.}, 2014). These observations are in concordance to the "multiple-hit" hypothesis that states that cancers do not arise from single mutagenic events, but are the consequence of sequential assaults on the genome that activate more than one oncogenic pathway. These multiple genetic hits allow cells to adopt traits to form cancer, which is coined by Bob Weinberg and Douglas Hanahan as the "Hallmarks of Cancer" (Hanahan and Weinberg, 2011; Hanahan, 2022). The traits include suppression of apoptosis, active proliferation, altered metabolism, changes in cell identity, and forming blood-vessels to supply tumor cells with adequate resources for growth(Hanahan, 2022). BRAF(V600E) expression alone is sufficient to activate many of the hallmarks of cancer but additional pathways need to be activated, or suppresed, to develop to malignant tumors. Great effort has elucidated the proto-oncogenes that drive cancer formation when mutated, such as \emph{MYC}, \emph{RAS}, \emph{PI3K}, and others. Yet we still do not fully understand the mechanistic details that determine how genes cooperate to overide tumor suppressive mechanisms and drive cancer progression.

@@

The importance of p53 is best demonstrated in people, and in mice, that have just one mutated copy of P53. In humans with the familial- inherited Li-Fraumeni syndrome, in which they have one or more mutations in p53 are almost certain to develop cancer early on and throughout their life(Li \emph{et al.}, 1988). Similarly, mice lacking two function P53 genes are prone to leukemias and lymphomas and have a shortened lifespan due to cancer incidence(Donehower \emph{et al.}, 1992). On the other hand, organisms such as the african and asian elephants, which have an abnormally low level of cancer incidence, have been shown to higher multiple copies of \emph{TP53} in their genomes(Abegglen \emph{et al.}, 2015).

The most well-known cellular functions of P53 include its ability to induce cell-cycle arrest in response to DNA damage(el-Deiry \emph{et al.}, 1993). Many stress signals, including oncogene activation has been shown to stimulate a reversible or irreversible cell cycle arrest(Zhu \emph{et al.}, 1998; McNeal \emph{et al.}, 2021). However, the strength of P53 induced cell cycle arrest or apoptosis is likely cell type- and cellular stress- specific and is a active area of p53 research.

@@

P53 was originally thought to be an oncogene as P53 is accumulated in lung cancer, which is not common in normal tissues(Iggo \emph{et al.}, 1990; Bartek \emph{et al.}, 1990). Moreover, ectopic expression of a P53 cDNA was found to aide in the transformation of primary cells induced by RAS. However, early studies erroneously used mutated P53 instead of wild-type leading to the misclassification of P53 as an oncogene. We now know P53 suppresses cell growth and transformation(Levine and Oren, 2009).

Often times, P53 is mutated in the DNA-binding domain at sites commonly referred to as "hotspots" due to there extraordinary frequency. These hotspot mutations poison the ability of p53 to bind DNA therefore blocking its function. Interestingly, since one mutated P53 can disrupt the entire P53 tetramer complex, a single mutation can exert dominant-negative effects by inhibiting the normal tumor suppressive functions of P53. Although one mutant P53 allele is enough to compromised transcriptional activity, there is still selective pressure to lose the other wild -type copy. Loss of hemizygosity implies there is still residual tumor suppression in the presence of a wild-type P53 allele(Muzumdar \emph{et al.}, 2016).

Early on in P53 studies, there were reports of certain P53 mutations having a "gain-of-function" (GOF) effect. For example, Li-Fraumeni patients with certain missense mutations in P53 would develop tumors earlier than Li-Fraumeni patients with loss-of-function (LOF) mutations (Bougeard \emph{et al.}, 2008). Furthermore, there were experimental cell biology studies that would express mutant-p53 in P53-null cells and demonstrate enhanced tumorigenic potential (Levine, 2020). Further evidence for GOF mutant-P53 in mice indicated that missense mutant P53 induced different cancer types and enhanced metastasis than LOF P53 (Olive \emph{et al.}, 2004). Furthermore, many groups have shown mutant-p53 can alter signal transduction affecting chemoresistance and altering metabolism.

The cellular mechanisms that GOF P53 are involved in are also well-characterized but context-dependent. For example, in Pancreatic cancers driven by KRAS(G12D) Mutant P3 interacts with CREB to induce FOXA1 transcription which enhances Beta-Catenin signaling to augment liver metastasis (Kim \emph{et al.}, 2021).

@@

Next generation sequencing has allowed for the unbiased identification of genes that are mutated in human cancer (2014; Hoadley \emph{et al.}, 2014; Kandoth \emph{et al.}, 2013).
The wealth of knowledge gained from such large-scale sequencing projects has revealed actionable mutationas that can be targeted with small molecule inhibitors, such as dabrafenib and trametinib.
However, Identificaton of possible targets is not enough, as many cancers with similarly mutated oncogenes, such as BRAF\textsuperscript{V600E}, respond differentially to BRAF\textsuperscript{V600E}-targeed therapy.
For example, mealanoma patients with BRAF\textsuperscript{V600E} mutations respond well to BRAF\textsuperscript{V600E}-targeted therapy whereas colorectal cancer patients with BRAF\textsuperscript{V600E} mutations do not respond(Corcoran \emph{et al.}, 2012).
The disconnect between putative driver oncogenes identified in human cancers and response to targeted therapy is a problem that can be solved with rigorous pre-clinical experimentation, such as in GEMMs.

Genomic sequencing of human tumors has also been crucial insight into how cancers form and what signaling nodes might be important for progressive stages of cancer evolution.
For example, pancreatic cancers have a very stereotypical histiological progression that is brought about by known mutational activation.
Specifically, mutations in \emph{KRAS} initiate tumorigenesis and loss of \emph{TP53} function and \emph{SMAD3} are associated with cancer progression(Orth \emph{et al.}, 2019).
However, this is more difficult to do in lung adenocarcinoma because of the high mutational burden often due to the consumption of cigarettes(Alexandrov \emph{et al.}, 2016; Yoshida \emph{et al.}, 2020).

@@

Recently, Monte Winslow and his lab at Stanford University has developed \textbf{Tu}mor- \textbf{Ba}rcode \textbf{Seq}uencing (Tuba-Seq), a rapid, multiplexed, quantitative assay for analyzing the contribution of tumor-suppressor loss \emph{in vivo}.
This technique works by delivering a uniquely barcoded lentiviral endoing Cre recombinase, to activate Cre-mediated oncogenes, and a guide RNA, to inactivate a single tumor suppressor gene.
Therefore, each tumor will be driven by the same oncogene, but will have different tumor suppressors inactivated.
After a period of time (6-24 weeks) tissue is harvested for whole-lung DNA extraction and targeted DNA-sequencing is performed to quantify relative tumor size to spike-in controls(Rogers \emph{et al.}, 2017).
Therefore, contribution of a single tumor suppressor can be quantified in single animal.

BRAF\textsuperscript{V600E} has previously been shown to be a potent driver of lung tumorigenesis(Dankort \emph{et al.}, 2007).
However, tumor fails to progress past adenoma-like tumors.
Given the theory that BRAF-\textsuperscript{V600E} induces growth arrest after an initial burst of neoplastic growth, it would be of interest to know if all tumors grow to a predetermined and absolute size, or if there is a spectrum of tumor sizes that are determined by extrinsic factors such as proximity to blood vessels or certain stromal cells.
Therfore, we used Tuba-Seq combined with CRISPR/CAS9 genome editing to explore the contribution of 11 tumor suppressor genes and 5 controls in the context of our BRAF \textsuperscript{V600E}- driven lung tumor models.

@@

Adenoviral vectors and lentiviral vectors introduced in the lungs of GEMMs offer temporal control and can be titrated to acheive an optimal level of tumor numerosity.
However, only lentiviral vectors can be used to barcode tumors because they integrate into the genome while adenoviruses do not(DuPage \emph{et al.}, 2009; Rogers \emph{et al.}, 2017).
Together with CRISPR/CAS9 gene editing, lentiviruses have been used to analyzed many genetic interactions of oncogenic \emph{Kras} driven lung adenocarcinoma(Winters \emph{et al.}, 2017; Rogers \emph{et al.}, 2017).
To enable simultaneous actvation of BRAF\textsuperscript{V600E} and loss of tumor supressor genes, we bred \emph{Braf\textsuperscript{CAT}}; \emph{H11b\textsuperscript{LSL-CAS9/+}} mice.
To determine the size of BRAF \textsuperscript{V600E} tumors and the contribution of 11 tumor suppressor genes (\emph{Trp53, Rb1, Cdkn2a, Apc, Rbm10, Stk11, Setd2, Arid1a, Atm, Keap1, Smad4)} on BRAF \textsuperscript{V600E} lung tumorigenesis, we obtained tittered barcoded-lentiviral plasmids that encode Cre-recombinase as well as a sgRNA that inactivates one of eleven tumor suppressors common in lung adenocarcinoma.
All genes tested are putative tumor suppressor genes commonly lost in human lung cancer patients. Each lentiviral solution contained \textasciitilde1000 individually barcoded lentiviruses.
We infected both \emph{LSL-Cas9} and \emph{CAS9} deficient mice in order to determine the contribution of genetic manipulation and/or deletion of the targeted gene.
Mice meant to be analyzed after 8 weeks received a viral titer of 1x10\^{}5 while mice mean to be analyzed after 16 weeks received a viral titer of 1x10\^{}4. After euthanasia, lungs were collected for genomic DNA extraction, and barcode amplification followed by preperation of sequencing libraries.
Before genomic extraction and library preparation, a known quantity of known-barcoded cells was added in order to benchmark sequencing reads to a known cell number(figure).

@@

Many GOF phenotypes exhibited by mutant-p53 have centered around drug resistant mechanisms. Previous studies have shown that mutant-p53 can confer resistance to pathway targeted therapy and chemotherapies. Our studies highlight that p53\textsuperscript{R172H} and p53\textsuperscript{R245W} can confer resistance to BRAF-targeted therapy. Here, we show p53-R172H and p53-R245W cell lines are resistant to BRAF(V600E) inhibition and less sensitivity to the combination of BRAF+MEK inhibition. In pancreatic cancer models with LSL-p53-R172H expression, mutant-p53 regulates PDGFR-signaling to promote an invasive phenotype that is sensitive to Imatinib(Weissmueller \emph{et al.}, 2014). Consistent with these findings, our transcriptional analysis and immunoblotting indicate p53-R172H and p53-R245W cells express PDGF-ligands and exhibit PDGFR-pathway activation.

>>>>

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/1af40d40f8291eb85788543d7614044c788929f1/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/1af40d40f8291eb85788543d7614044c788929f1}{scherzerthesis/thesis@1af40d4}
on March 30, 2022.

@@

\textbf{NSCLC} Lung cancer is a global hazard on human health. The historic consumption of cigarette smoke has caused lung cancer to be the most common and deadly form of cancer. Fortunately, smoking-rates have decreased in the past 30 years with lung cancer rates also on the decline. However, even if cigarette consumption were eradicated, lung cancer would still significantly affect human health. Indeedm it is estimated that approximately 25\% of all lung cancer cases arise in never smokers (Warren and Cummings, 2013). Secondary causes of lung cancer include Radon exposure, high-heat cooking, family history, air pollution (Malhotra et al., 2016). An individual's risk of lung cancer also increases with age due to decreases in the efficiency of tumor supressive mechanisms such as tumor immunity and apoptosis(de Groot and Munden, 2012).

Lung cancer can be broadly classified into two major histological categories that are named based on how the malignant cells look under a microscope: Non-small Cell Lung Cancer (NSCLC) and Small Cell Lung Cancer (SCLC). Within NSCLC, tumors are further characterized based on histological structures, anatomical location, cell-of-origin, as well as genetic drivers(Herbst et al., 2008). For example, lung adenocarcinoma is the most common subtype and is characterized by glandular and papillary structures while squamous cell lung carcinoma is characterized by keratin-pearls(Nikitin et al., 2004). Moreover, adenocarcinomas arise from alveolar-type-2 (AT2) cells in the distal lung and alveoli and are typified by genomic alterations throughout the MAPK pathway, such as EGFR, KRAS, BRAF, PI3K (Ferone et al., 2020; Jackson et al., 2001). On the other hand, squamous cells carcinomas likely arise from basal cells and typically harbor SOX2 gene amplification (Mollaoglu et al., 2018; Mukhopadhyay and Oliver, 2014).

The 5-year survival rate for patients with lung cancer was 25\% in 2020 but varies depending on stage of disease at time of diagnosis (Siegel et al., 2021).Like other solid cancers, lung cancers are most deadly after primary tumors have metastaized to other organ systems like the liver, kidney, bone, or nervous system (Matsunaga and Shida, 1971). Lung cancer patients with no local or distant metastasis typically have a 5-year survival percentage of approximately 40\% compared to approximately 5\% with distant metastasis{[}Woodard et al.; 27740970{]}. Ultimately, the cause of death of many lung cancer patients include primary lung tumor burden, infection, pulmonary embolism or hemorrhage, or complications of metastatic dissemination(Nichols et al., 2012). Although patient survival has improved due to improved targeted and immune therapies, the still poor prognosis reflects a need to better understand the molecular mechanisms underlying lung adenocarcinoma initiation, progression, maintenance and response to targeted- or immune-therapy.

The current standard of care for Lung adenocarcinoma patients are often a combination of several conventional chemotherapeutics including Cisplatin, Carboplatin, Paclitaxel, or Pemetrexed(Ghimessy et al., 2020). Fortunately for patients who have been identified to have EGFR, ALK, ROS, TRK, or BRAF (V600E) genetic alternations are often offered pathway-targeted therapies(Arteaga and Engelman, 2014; Planchard et al., 2016). As of late, many patients receive immunotherapy first, followed by pathway-targeted therapy after progression on immunotherapy. Furthermore, in the last seven years, exciting therapeutic options for \emph{KRAS} mutant lung cancer have been developed and tested, such as a combination of MEK-inhibition and autophagy-inhibtion, and direct inhibitors of \emph{KRAS} G12c and \emph{KRAS} G12D (Kinsey et al., 2019; Reita et al., 2022; Zheng et al., 2022)

@@

The importance of this pathway is reflected in the strong anti-tumor responses that are seen when small molecule inhibitors are used to target various molecules involved in the MAPK pathway. For example, Dabrafenib and Trametinib, inhibitors of BRAF and MEK kinases, respectfully, are used in patients that harbor BRAF(V600E) melanomas and lung cancers(Planchard et al., 2021; Robert et al., 2014). Although there is strong pre-clinical and clinical evidence that BRAF+MEK blockade can lead to tumor regression, not all patients respond to treatment(Tabbò et al., 2021). Therefore, to see a complete response, multiple pathways need to be targeted, such as PI3K or autophagy (Deuker et al., 2014; Kinsey et al., 2019; Truong et al., 2020).

@@

\textbf{Genetically-engineered mouse models} Genetically-Engineered mouse models (GEMMs) of human cancer have allowed for the reproducible analysis of cancer biology driven by specific mutations frequently found in cancer patients. These models are helpful in understanding the molecular mechanisms that drive tumor formation and testing tumor responses to targeted or immune therapies. Early mouse models were not engineered but instead selected inbred mouse strains that were natural hosts of tumor transplantation(Strong, 1978). However, these models suffer from long latencies, incomplete pentrance, and often not relevant to human cancers. Therefore, carcinogen-induced models such as UV, Urethane, radiation, cigarette consumption, or MNU treatment models were developed to ressemble carinogene-induced human cancers(Kemp et al., 1994; Westcott et al., 2014).After techniques were developed to genetically manipulate embryonic stem-cells came the ability to engineer mice with specific mutations that could drive tumor formation. Since genes involved in cancer are also important for normal mouse development, conditional gene manipulation was developed to allow for temporal control.

The most commonly used conditional gene expression system relies on inducible transcription factors, such as tetracycline-dependent regulatory system, where the treatment of mice with doxycycline toggles gene transcriptial levels of transgenes. More recently, Cre-Lox recombination strategies have been utilized to get more cell type specific control. In this system, genes of interested are constructed to contain LoxP-sites that flank a particular genetic element and upon delivery of Cre-recombinase the DNA sequence in between LoxP sites are excised. With this approach, one can knockout a gene by engineering LoxP sites around critical exons, or one could induce gene transcription by flanking a strong stop signal such as a polyadenylation sequence upstream of a gene-of-interest sequence such as KRAS(G12D)(Jackson et al., 2001).

@@

Our lab has previously developed a conditional mouse model of BRAF (V600E) human lung cancers(Dankort et al., 2007; Shai et al., 2015; van Veen et al., 2019). In the \emph{Braf(CAT)} model, normal BRAF is expressed from a conditional allele prior to Cre-mediated recombination . LoxP sites flank human \emph{BRAF} cDNA encoding normal \emph{BRAF} exons 15-18. Downstream of the LoxP sites is the mutant exon 15 that encodes the murine equivalent of the T1799A mutation that gives rise to the BRAF (V600E) onco-protein. Downstream of the mutant exon there is a P2A element and a CAAX-tagged \emph{TdTomato}. Therefore, after Cre-mediated recombination, the BRAF (V600E) oncoprotein and fluorophore TdTomato is expressed at normal levels(van Veen et al., 2019).

\textbf{BRAF (V600E) Tumorigenesis and Cooperating Events for Progression} Mutational activation of the MAPK pathway forms lung tumors in mice. specifically, expression of BRAF (V600E) in the Surfactant-Protein C expression cells (alveolar type 2 pneumocytes) elicits benign tumor formation that fails to progress to lung adenocarcinoma. It is hypothesized that the cell cycle arrest is dependent on the tumor suppressors P53 and CDKN2A. Loss of either P53 or CDKN2A allows tumors to progress to malignant and deadly lung adenocarcinoma. Also, activation of WNT signaling and PI3K signaling similarly allows BRAF (V600E) adenomas to progress to adenocarcinoma (Green et al., 2015; Juan et al., 2014; van Veen et al., 2019). These observations are in concordance to the "multiple-hit" hypothesis that states that cancers do not arise from single mutagenic events, but are the consequence of sequential assaults on the genome that activate more than one oncogenic pathway. These multiple genetic hits allow cells to adopt traits to form cancer, which is coined by Bob Weinberg and Douglas Hanahan as the "Hallmarks of Cancer" (Hanahan, 2022; Hanahan and Weinberg, 2011). The traits include suppression of apoptosis, active proliferation, altered metabolism, changes in cell identity, and forming blood-vessels to supply tumor cells with adequate resources for growth(Hanahan, 2022). BRAF(V600E) expression alone is sufficient to activate many of the hallmarks of cancer but additional pathways need to be activated, or suppresed, to develop to malignant tumors. Great effort has elucidated the proto-oncogenes that drive cancer formation when mutated, such as \emph{MYC}, \emph{RAS}, \emph{PI3K}, and others. Yet we still do not fully understand the mechanistic details that determine how genes cooperate to overide tumor suppressive mechanisms and drive cancer progression.

@@

The importance of p53 is best demonstrated in people, and in mice, that have just one mutated copy of P53. In humans with the familial- inherited Li-Fraumeni syndrome, in which they have one or more mutations in p53 are almost certain to develop cancer early on and throughout their life(Li et al., 1988). Similarly, mice lacking two function P53 genes are prone to leukemias and lymphomas and have a shortened lifespan due to cancer incidence(Donehower et al., 1992). On the other hand, organisms such as the african and asian elephants, which have an abnormally low level of cancer incidence, have been shown to higher multiple copies of \emph{TP53} in their genomes(Abegglen et al., 2015).

The most well-known cellular functions of P53 include its ability to induce cell-cycle arrest in response to DNA damage(el-Deiry et al., 1993). Many stress signals, including oncogene activation has been shown to stimulate a reversible or irreversible cell cycle arrest(McNeal et al., 2021; Zhu et al., 1998). However, the strength of P53 induced cell cycle arrest or apoptosis is likely cell type- and cellular stress- specific and is a active area of p53 research.

@@

P53 was originally thought to be an oncogene as P53 is accumulated in lung cancer, which is not common in normal tissues(Bartek et al., 1990; Iggo et al., 1990). Moreover, ectopic expression of a P53 cDNA was found to aide in the transformation of primary cells induced by RAS. However, early studies erroneously used mutated P53 instead of wild-type leading to the misclassification of P53 as an oncogene. We now know P53 suppresses cell growth and transformation(Levine and Oren, 2009).

Often times, P53 is mutated in the DNA-binding domain at sites commonly referred to as "hotspots" due to there extraordinary frequency. These hotspot mutations poison the ability of p53 to bind DNA therefore blocking its function. Interestingly, since one mutated P53 can disrupt the entire P53 tetramer complex, a single mutation can exert dominant-negative effects by inhibiting the normal tumor suppressive functions of P53. Although one mutant P53 allele is enough to compromised transcriptional activity, there is still selective pressure to lose the other wild -type copy. Loss of hemizygosity implies there is still residual tumor suppression in the presence of a wild-type P53 allele(Muzumdar et al., 2016).

Early on in P53 studies, there were reports of certain P53 mutations having a "gain-of-function" (GOF) effect. For example, Li-Fraumeni patients with certain missense mutations in P53 would develop tumors earlier than Li-Fraumeni patients with loss-of-function (LOF) mutations (Bougeard et al., 2008). Furthermore, there were experimental cell biology studies that would express mutant-p53 in P53-null cells and demonstrate enhanced tumorigenic potential (Levine, 2020). Further evidence for GOF mutant-P53 in mice indicated that missense mutant P53 induced different cancer types and enhanced metastasis than LOF P53 (Olive et al., 2004). Furthermore, many groups have shown mutant-p53 can alter signal transduction affecting chemoresistance and altering metabolism.

The cellular mechanisms that GOF P53 are involved in are also well-characterized but context-dependent. For example, in Pancreatic cancers driven by KRAS(G12D) Mutant P3 interacts with CREB to induce FOXA1 transcription which enhances Beta-Catenin signaling to augment liver metastasis (Kim et al., 2021).

@@

Next generation sequencing has allowed for the unbiased identification of genes that are mutated in human cancer (2014; Hoadley et al., 2014; Kandoth et al., 2013).
The wealth of knowledge gained from such large-scale sequencing projects has revealed actionable mutationas that can be targeted with small molecule inhibitors, such as dabrafenib and trametinib.
However, Identificaton of possible targets is not enough, as many cancers with similarly mutated oncogenes, such as BRAF\textsuperscript{V600E}, respond differentially to BRAF\textsuperscript{V600E}-targeed therapy.
For example, mealanoma patients with BRAF\textsuperscript{V600E} mutations respond well to BRAF\textsuperscript{V600E}-targeted therapy whereas colorectal cancer patients with BRAF\textsuperscript{V600E} mutations do not respond(Corcoran et al., 2012).
The disconnect between putative driver oncogenes identified in human cancers and response to targeted therapy is a problem that can be solved with rigorous pre-clinical experimentation, such as in GEMMs.

Genomic sequencing of human tumors has also been crucial insight into how cancers form and what signaling nodes might be important for progressive stages of cancer evolution.
For example, pancreatic cancers have a very stereotypical histiological progression that is brought about by known mutational activation.
Specifically, mutations in \emph{KRAS} initiate tumorigenesis and loss of \emph{TP53} function and \emph{SMAD3} are associated with cancer progression(Orth et al., 2019).
However, this is more difficult to do in lung adenocarcinoma because of the high mutational burden often due to the consumption of cigarettes(Alexandrov et al., 2016; Yoshida et al., 2020).

@@

Recently, Monte Winslow and his lab at Stanford University has developed \textbf{Tu}mor- \textbf{Ba}rcode \textbf{Seq}uencing (Tuba-Seq), a rapid, multiplexed, quantitative assay for analyzing the contribution of tumor-suppressor loss \emph{in vivo}.
This technique works by delivering a uniquely barcoded lentiviral endoing Cre recombinase, to activate Cre-mediated oncogenes, and a guide RNA, to inactivate a single tumor suppressor gene.
Therefore, each tumor will be driven by the same oncogene, but will have different tumor suppressors inactivated.
After a period of time (6-24 weeks) tissue is harvested for whole-lung DNA extraction and targeted DNA-sequencing is performed to quantify relative tumor size to spike-in controls(Rogers et al., 2017).
Therefore, contribution of a single tumor suppressor can be quantified in single animal.

BRAF\textsuperscript{V600E} has previously been shown to be a potent driver of lung tumorigenesis(Dankort et al., 2007).
However, tumor fails to progress past adenoma-like tumors.
Given the theory that BRAF-\textsuperscript{V600E} induces growth arrest after an initial burst of neoplastic growth, it would be of interest to know if all tumors grow to a predetermined and absolute size, or if there is a spectrum of tumor sizes that are determined by extrinsic factors such as proximity to blood vessels or certain stromal cells.
Therfore, we used Tuba-Seq combined with CRISPR/CAS9 genome editing to explore the contribution of 11 tumor suppressor genes and 5 controls in the context of our BRAF \textsuperscript{V600E}- driven lung tumor models.

@@

Adenoviral vectors and lentiviral vectors introduced in the lungs of GEMMs offer temporal control and can be titrated to acheive an optimal level of tumor numerosity.
However, only lentiviral vectors can be used to barcode tumors because they integrate into the genome while adenoviruses do not(DuPage et al., 2009; Rogers et al., 2017).
Together with CRISPR/CAS9 gene editing, lentiviruses have been used to analyzed many genetic interactions of oncogenic \emph{Kras} driven lung adenocarcinoma(Rogers et al., 2017; Winters et al., 2017).
To enable simultaneous actvation of BRAF\textsuperscript{V600E} and loss of tumor supressor genes, we bred \emph{Braf\textsuperscript{CAT}}; \emph{H11b\textsuperscript{LSL-CAS9/+}} mice.
To determine the size of BRAF \textsuperscript{V600E} tumors and the contribution of 11 tumor suppressor genes (\emph{Trp53, Rb1, Cdkn2a, Apc, Rbm10, Stk11, Setd2, Arid1a, Atm, Keap1, Smad4)} on BRAF \textsuperscript{V600E} lung tumorigenesis, we obtained tittered barcoded-lentiviral plasmids that encode Cre-recombinase as well as a sgRNA that inactivates one of eleven tumor suppressors common in lung adenocarcinoma.
All genes tested are putative tumor suppressor genes commonly lost in human lung cancer patients. Each lentiviral solution contained \textasciitilde1000 individually barcoded lentiviruses.
We infected both \emph{LSL-Cas9} and \emph{CAS9} deficient mice in order to determine the contribution of genetic manipulation and/or deletion of the targeted gene.
Mice meant to be analyzed after 8 weeks received a viral titer of 1x10\^{}5 while mice mean to be analyzed after 16 weeks received a viral titer of 1x10\^{}4. After euthanasia, lungs were collected for genomic DNA extraction, and barcode amplification followed by preperation of sequencing libraries.
Before genomic extraction and library preparation, a known quantity of known-barcoded cells was added in order to benchmark sequencing reads to a known cell number(figure).

@@

Many GOF phenotypes exhibited by mutant-p53 have centered around drug resistant mechanisms. Previous studies have shown that mutant-p53 can confer resistance to pathway targeted therapy and chemotherapies. Our studies highlight that p53\textsuperscript{R172H} and p53\textsuperscript{R245W} can confer resistance to BRAF-targeted therapy. Here, we show p53-R172H and p53-R245W cell lines are resistant to BRAF(V600E) inhibition and less sensitivity to the combination of BRAF+MEK inhibition. In pancreatic cancer models with LSL-p53-R172H expression, mutant-p53 regulates PDGFR-signaling to promote an invasive phenotype that is sensitive to Imatinib(Weissmueller et al., 2014). Consistent with these findings, our transcriptional analysis and immunoblotting indicate p53-R172H and p53-R245W cells express PDGF-ligands and exhibit PDGFR-pathway activation.
